Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-22 doc/thesis/intro.tex
Stuff to work on for dynamics

A caveat of estimating transmission trees in this manner is that the
correspondence between the topologies of the viral phylogeny and transmission
tree is not necessarily exact~\autocite{ypma2013relating, romero2014timing}.
Due to intra-host diversity, the viral strain which is transmitted may have
split from another lineage within the donor long before the transmission event
occurred. Hence, the branching point in the viral phylogeny may be much earlier
than that in the transmission tree. Another possibility is that one host
transmitted to two or more recipients, but the lineages they each received
originated within the donor host in a different order than that in which the
transmissions occurred.  In this case, the topology of the transmission tree
and the viral phylogeny will be mismatched. In practice, this discordance has
not proven an insurmountable problem: for example,
\textcite{leitner1996accurate, paraskevis2004phylogenetic} were able to
accurately recover a known transmission tree using a viral phylogeny. The
problem of accurately estimating transmission trees is an ongoing area of
research~\autocite{cottam2008integrating, jombart2011reconstructing,
ypma2012unravelling, didelot2014bayesian}.

@@

Epidemiological models most often assume some form of contact homogeneity. The
simplest models, such as the \gls{SIR} model, assume a completely homogeneously
mixed population, where every pair of contacts is equally likely. More
sophisticated models partition the population into groups with different
contact rates between and among each group. However, these models still assume
that every possible contact between a member of group $i$ and a member of group
$j$ is equally likely. This assumption is clearly unrealistic for the majority
of human communities, and can lead to significant errors in predicted epidemic
trajectories when there is substantial heterogeneity
present~\autocite{bansal2007individual, volz2007susceptible}. Contact networks
provide a way to relax this assumption by representing individuals and their
contacts explicitly. It is important to note that, although panmixia is an
unrealistic modelling assumption, it has not proven a substantial hurdle to
epidemic modelling in practice~\autocite{anderson1992infectious}. Using this
assumption, researchers have been able to derive estimates of the transmission
rate and the basic reproductive number of various outbreaks, which have agreed
with values obtained by on-the-ground data collection. Therefore, if one is
interested only in these population-level variables, the additional complexity
of contact network models may not be warranted. Rather, these models are most
useful when we are interested in properties of the network itself, such as
centrality, structural balance, and
transitivity~\autocite{wasserman1994social}.

@@

We define a \defn{Markov kernel} as the continuous analogue of the transition
matrix in a finite-state Markov model.  For some spaces $X$ and $Y$, $K \maps X
\times Y \to [0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. Note that a Markov kernels have nothing to
do with the kernel functions defined in \cref{subsec:treeshape}, other than
sharing a name (the word ``kernel'' is ubiquitous in mathematics).

@@

TODO

>>>>

A caveat of estimating transmission trees in this manner is that the
correspondence between the topologies of the viral phylogeny and transmission
tree is not necessarily exact~\autocite{ypma2013relating, romero2014timing}.
Due to intra-host diversity, the viral strain which is transmitted may have
split from another lineage within the donor long before the transmission event
occurred. Hence, the branching point in the viral phylogeny may be much earlier
than that in the transmission tree. Another possibility is that one host
transmitted to two or more recipients, but the lineages they each received
originated within the donor host in a different order than that in which the
transmissions occurred.  In this case, the topology of the transmission tree
and the viral phylogeny will be mismatched. In practice, this discordance has
not proven an insurmountable problem: for example,
\textcite{leitner1996accurate, paraskevis2004phylogenetic} were able to
accurately recover a known transmission tree using a viral phylogeny. The
problem of accurately estimating transmission trees is an ongoing area of
research~\autocite{cottam2008integrating, jombart2011reconstructing,
ypma2012unravelling, morelli2012bayesian, didelot2014bayesian}.

@@

Epidemiological models most often assume some form of contact homogeneity. The
simplest models, such as the \gls{SIR} model, assume a completely homogeneously
mixed population, where every pair of contacts is equally likely. More
sophisticated models partition the population into groups with different
contact rates between and among each group. However, these models still assume
that every possible contact between a member of group $i$ and a member of group
$j$ is equally likely. This assumption is clearly unrealistic for the majority
of human communities, and can lead to significant errors in predicted epidemic
trajectories when there is substantial heterogeneity
present~\autocite{bansal2007individual, volz2007susceptible,
rolls2015simulation}. Contact networks provide a way to relax this assumption
by representing individuals and their contacts explicitly. It is important to
note that, although panmixia is an unrealistic modelling assumption, it has not
proven a substantial hurdle to epidemic modelling in
practice~\autocite{anderson1992infectious}. Using this assumption, researchers
have been able to derive estimates of the transmission rate and the basic
reproductive number of various outbreaks, which have agreed with values
obtained by on-the-ground data collection. Therefore, if one is interested only
in these population-level variables, the additional complexity of contact
network models may not be warranted. Rather, these models are most useful when
we are interested in properties of the network itself, such as centrality,
structural balance, and transitivity~\autocite{wasserman1994social}.

@@

We define a \defn{Markov kernel} as the continuous analogue of the transition
matrix in a finite-state Markov model.  For some spaces $X$ and $Y$, $K \maps X
\times Y \to [0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. A more rigorous definition can be found in
\eg \autocite{kallenberg2006foundations}. Note that a Markov kernels have
nothing to do with the kernel functions defined in \cref{subsec:treeshape},
other than sharing a name (the word ``kernel'' is ubiquitous in mathematics).

@@

### 2016-04-22 doc/thesis/methods.tex
Stuff to work on for dynamics

>>>>

Two further experiments were performed to address potential sources of error.
To evaluate the effect of model misspecification in the case of heterogeneity
among nodes, we generated a network where half the nodes were attached with
power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
investigate the effects of potential sampling
bias~\autocite{karcher2016quantifying}, we simulated a transmission tree where
the tips were sampled in a peer-driven fashion, rather than at random. That is,
the probability to sample a node was twice as high if any of that node's
network peers had already been sampled. The parameters of this network were $N$
= 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.

### 2016-04-22 doc/thesis/objective.tex
Stuff to work on for dynamics

The spread of a disease is most often modelled by assuming either a
homogeneously mixed population~\autocite{hamer1906milroy,
kermack1927contribution}, or a population divided into a small number of
homogeneously mixed groups~\autocite{rushton1955deterministic}. This
assumption, also called the ``law of mass action'' or \defn{panmixia}, implies
that any two individuals in the same compartment are equally likely to come
into contact causing transmission. Although this provides a reasonable
approximation in many cases, the error introduced by assuming a panmictic
population can be substantial when significant contact heterogeneity exists in
the underlying population. Contact network models provide an alternative to
compartmental models which do not require the assumption of panmixia. In
addition to more accurate predictions, the parameters of the networks
themselves may be of interest from a public health perspective. For example,
certain vaccination strategies may be more or less effective in curtailing an
epidemic depending on the underlying network's degree distribution.
Phylodynamic methods have been used to fit many different types of model to
phylogenetic data, but as far as we know, no methods have yet been developed to
fit contact network models. The primary objective of this work is to develop
such a method.

@@

Calculating the likelihood of the parameters of a contact network models seems
likely to be an intractable problem, which would imply that these models are
amenable to neither \gls{ML} nor Bayesian inference. We not proven this is the
case, but some intuition can be provided by examining the process involved in
the likelihood calculation. Consider a contact network model with parameters
$\theta$, and an observed transmission tree $T$ with $n$ tips.  In general, we
do not know the labels of the internal nodes of $T$, only the labels of its
tips. To fit this model using likelihood-based methods, we must calculate the
likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$ be the set of
all possible contact networks, and $\Nu$ be the set of all possible labellings
of the internal nodes of $T$. We can write the likelihood as
\begin{align}
\begin{split}
  \label{eq:netlik}
  \Pr(T \mid \theta)
    &= \sum_{\nu \in \Nu} \Pr(T, \nu \mid \theta) \\
    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta) \\
    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G) \Pr(G \mid \theta),
\end{split}
\end{align}
the last equality following from the fact that $T$ and $\nu$ depend only on
$G$, not on $\theta$. Although $\Pr(T, \nu \mid G)$ and $\Pr(G \mid \theta)$
may individually be straightforward to calculate, the number of possible
directed graphs on $N$ nodes is $2^{N(N-1)}$, larger if the nodes and edges in
the graph may have different labels or attributes. Hence, the number of terms
in the sum is at least exponential in $n$, as there must be at least $n$ nodes
in the network. In addition, \cref{eq:netlik} assumes that $T$ is complete,
meaning that all infected individuals were sampled. This is rarely the case in
practice - most often, the observed tree is a subsampled version of the true
tree. In this case, the likelihood calculation becomes even more complex,
because we must also sum over all possible complete trees.

Depending on the network model studied, it is possible that \cref{eq:netlik}
could be simplified into a tractable expression. However, a simpler alternative
to likelihood-based methods, which would apply to any network model, is
provided by \gls{ABC}. All of the ingredients required to apply \gls{ABC} to
this problem are readily available. Simulating networks is straightforward
under a variety of models. Epidemics on those networks, and the corresponding
transmission trees, can also be easily simulated. As mentioned above, contact
networks can profoundly affect transmission tree shape, and those shapes can be
compared using a highly informative similarity measure. \Gls{SMC} has several
advantages over other algorithms for \gls{ABC}~\autocite{mckinley2009inference},
including a recently-developed adaptive algorithm requiring minimal tuning on
the part of the user~\autocite{del2012adaptive}. In summary, our method to
infer contact network parameters will combine the following: stochastic
simulation of epidemics on networks, the tree kernel, and adaptive
\gls{ABC}-\gls{SMC}. Since our distance measure is a kernel function, our
method is a type of kernel-\gls{ABC}. For ease of exposition, we will often use
the term ``kernel-\gls{ABC}'' to refer to our method specifically.

Empirical studies of sexual contact networks have found that these networks
tend to be scale-free, meaning that their degree distributions follow a power
law (although there has been some disagreement). Preferential attachment has
been postulated as a mechanism by which scale-free networks could be generated.
This makes the \gls{BA} model, one of the simplest preferential attachment
models, a natural choice to explore with our method. The second aim of this
work is to use simulations to investigate the parameters of the \gls{BA} model,
including whether they have a detectable impact on tree shape, and whether they
can be accurately recovered using kernel-\gls{ABC}.

Due to its high global prevalence and fast mutation rate, \gls{HIV} is one of
the most commonly-studied viruses in a phylodynamic context. Consequently, a
large volume of \gls{HIV} sequence data is publicly available, more than for
any other pathogen, and including sequences sampled from diverse geographic and
demographic contexts. Since \gls{HIV} is almost always spread through either
sexual contact or sharing of injection drug supplies, the contact networks
underlying \gls{HIV} epidemics are highly structured. Moreover, since no cure
yet exists, efforts to curtail the progression of an epidemic have relied on
preventing further transmissions through measures such as \gls{tasp} and
education leading to behaviour change. The effectiveness of this type of
intervention can vary significantly based on the underlying structure of the
network and the particular nodes to whom the intervention is targeted. Due to
this combination of data availability and potential public health impact,
\gls{HIV} is an obvious context in which our method could be applied.
Therefore, the third and final aim of this work is to apply kernel-\gls{ABC} to
fit the \gls{BA} model to existing \gls{HIV} outbreaks.

>>>>

\glsreset{ABC}

The spread of a disease is most often modelled by assuming either a
homogeneously mixed population~\autocite{hamer1906milroy,
kermack1927contribution}, or a population divided into a small number of
homogeneously mixed groups~\autocite{rushton1955deterministic}. This
assumption, also called the ``law of mass action'' or \defn{panmixia}, implies
that any two individuals in the same compartment are equally likely to come
into contact causing transmission. Although this provides a reasonable
approximation in many cases~\autocite{anderson1992infectious}, the error
introduced by assuming a panmictic population can be substantial when
significant contact heterogeneity exists in the underlying population. Contact
network models provide an alternative to compartmental models which do not
require the assumption of panmixia. In addition to more accurate predictions,
the parameters of the networks themselves may be of interest from a public
health perspective. For example, certain vaccination strategies may be more or
less effective in curtailing an epidemic depending on the underlying network's
degree distribution~\autocite{peng2013vaccination}. Phylodynamic methods have
been used to fit many different types of model to phylogenetic
data~\autocite{pybus2009evolutionary,volz2013viral}, but as far as we know, no
methods have yet been developed to fit contact network models. The primary
objective of this work is to develop such a method.

@@

Calculating the likelihood of the parameters of a contact network models seems
likely to be an intractable problem, which would imply that these models are
amenable to neither \gls{ML} nor Bayesian inference. We have not proven this is
the case, but some intuition can be provided by examining the process involved
in the likelihood calculation. Consider a contact network model with parameters
$\theta$, and an estimated transmission tree $T$ with $n$ tips. In general, we
do not know the labels of the internal nodes of $T$, only the labels of its
tips. To fit this model using likelihood-based methods, we must calculate the
likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$ be the set of
all possible contact networks, and $\Nu$ be the set of all possible labellings
of the internal nodes of $T$. We can write the likelihood as
\begin{align}
\begin{split}
  \label{eq:netlik}
  \Pr(T \mid \theta)
    &= \sum_{\nu \in \Nu} \Pr(T, \nu \mid \theta) \\
    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta) \\
    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G) \Pr(G \mid \theta),
\end{split}
\end{align}
the last equality following from the fact that $T$ and $\nu$ depend only on
$G$, not on $\theta$. Although $\Pr(T, \nu \mid G)$ and $\Pr(G \mid \theta)$
may individually be straightforward to calculate, the number of possible
directed graphs on $N$ nodes is $2^{N(N-1)}$, larger if the nodes and edges in
the graph may have different labels or attributes. Hence, the number of terms
in the sum is at least exponential in $n$, as there must be at least $n$ nodes
in the network. In addition, \cref{eq:netlik} assumes that $T$ is complete,
meaning that all infected individuals were sampled. This is rarely the case in
practice - most often, we only have access to a subset of the infected
individuals. In this case, the likelihood calculation becomes even more
complex, because we must also sum over all possible complete trees.

Depending on the network model studied, it is possible that \cref{eq:netlik}
could be simplified into a tractable expression. However, a simpler alternative
to likelihood-based methods, which would apply to any network model, is
provided by \gls{ABC}. All of the ingredients required to apply \gls{ABC} to
this problem are readily available. Simulating networks is straightforward
under a variety of models. Epidemics on those networks, and the corresponding
transmission trees, can also be easily simulated. As mentioned above, contact
networks can profoundly affect transmission tree shape, and those shapes can be
compared using a highly informative similarity measure called the ``tree
kernel''~\autocite{poon2013mapping}. \Gls{ABC} can be implemented with
\gls{SMC}, which has several advantages over other
algorithms~\autocite{mckinley2009inference}. A recently-developed adaptive
algorithm requiring minimal tuning on the part of the user makes \gls{SMC} an
even more attractive approach~\autocite{del2012adaptive}. In summary, our
method to infer contact network parameters will combine the following:
stochastic simulation of epidemics on networks, the tree kernel, and adaptive
\gls{ABC}-\gls{SMC}. Since our distance measure is a kernel function, our
method is a type of kernel-\gls{ABC}. For ease of exposition, we will often use
the term ``kernel-\gls{ABC}'' to refer to our method specifically.

Empirical studies of sexual contact networks have found that these networks
tend to be scale-free~\autocite{colgate1989risk, liljeros2001web,
schneeberger2004scale}, meaning that their degree distributions follow a power
law (although there has been some disagreement, see
\autocite{handcock2004likelihood, bansal2007individual}). Preferential
attachment has been postulated as a mechanism by which scale-free networks
could be generated~\autocite{barabasi1999emergence}. This makes the \gls{BA}
model, one of the simplest preferential attachment models, a natural choice to
explore with our method. The second aim of this work is to use simulations to
investigate the parameters of the \gls{BA} model, including whether they have a
detectable impact on tree shape, and whether they can be accurately recovered
using kernel-\gls{ABC}.

Due to its high global prevalence and fast mutation rate, \gls{HIV} is one of
the most commonly-studied viruses in a phylodynamic context. Consequently, a
large volume of \gls{HIV} sequence data is publicly available, more than for
any other pathogen, and including sequences sampled from diverse geographic and
demographic contexts. At the time of this writing, there were $635400$ HIV
sequences publicly available in GenBank, annotated with 172 distinct countries
of origin. Since \gls{HIV} is almost always spread through either sexual
contact or sharing of injection drug supplies, the contact networks underlying
\gls{HIV} epidemics are driven by social dynamics and are therefore likely to
be highly nonrandom. Moreover, since no cure yet exists, efforts to curtail the
progression of an epidemic have relied on preventing further transmissions
through measures such as \gls{tasp} and education leading to behaviour change.
The effectiveness of this type of intervention can vary significantly based on
the underlying structure of the network and the particular nodes to whom the
intervention is targeted. Due to this combination of data availability and
potential public health impact, \gls{HIV} is an obvious context in which our
method could be applied. Therefore, the third and final aim of this work is to
apply kernel-\gls{ABC} to fit the \gls{BA} model to existing \gls{HIV}
outbreaks.

### 2016-04-22 doc/thesis/preface.tex
Stuff to work on for dynamics

A version of chapter 2 has been submitted for publication in Virus Evolution
with the title ``Reconstructing network parameters from viral phylogenies.'' A
presentation entitled ``Phylodynamic inference of contact network parameters
with kernel-ABC'' was given based on chapter 2 at the 23rd HIV Dynamics and
Evolution meeting on April 25, in Woods Hole, Massachussets, USA.

>>>>

A version of chapter 2 has been submitted for publication in Virus Evolution
with the title ``Reconstructing network parameters from viral phylogenies.'' An
oral presentation entitled ``Phylodynamic inference of contact network
parameters with kernel-ABC'' was given based on chapter 2 at the 23rd HIV
Dynamics and Evolution meeting on April 25, in Woods Hole, Massachussets, USA.
A poster entitled ``Likelihood-free estimation of contact network parameters
from viral phylogenies'' was presented at the Intelligent Systems for Molecular
Biology meeting in Orlando, Florida, USA.

### 2016-04-22 doc/thesis/results.tex
Stuff to work on for dynamics

The average error for \gls{m} was
    1.31,
on a grid from 1 to 6; this error was positively correlated with increasing
\gls{m}
    (Spearman's $\rho$ = 0.25,
     $p = 6\!\times\!10^{-4}$).
This positive correlation was apparently due to the much lower error for
\gls{m} = 1 and 2 than for the other $m$ values 
    (mean errors 
    0.93 
    for $m \leq 2$ vs. 
    1.49 
    for $m > 2$, \cref{fig:gridm}).
The value $m = 1$ causes the network to take on a distinct shape relative to
higher \gls{m} values, namely a tree (\textit{i.e.} there are no cycles,
see \cref{subsec:treeshape}). The average error for \gls{N} was 
    2419,
on a grid from 1000 to 15000, and was positively correlated with the true value
of \gls{N}
    (Spearman's $\rho$ = 0.43,
    $p {<}10^{-5}$).

@@

We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
and found substantial heterogeneity among the parameter estimates
(\cref{fig:abchpd}). Two of the datasets~\autocite{niculescu2015recent,
wang2015targeting} had estimated $\alpha$ values near unity (MAP estimates
[95\% HPD] 
  1.06 
  [0.63 - 
   1.27]
and
  1 
  [0.41 -
   1.16] respectively). 
Another two datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated
values and wider HPD intervals
  (0.77 
  [0.01 - 
  1.03]
and
  0.66 
  [0.03 -
   0.84]). 
The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
and a very wide HPD interval
  (0.17 
  [0.04 -
   1.39]). 
For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
$I$ were below 2000, with narrow HPD intervals around two of the
datasets
  (\citeauthor{cuevas2009hiv}, 880 
  [290 -
   \ensuremath{1.51\times 10^{3}}];
   \citeauthor{niculescu2015recent}, 175
  [138 - 
   454])
and wider intervals around the other two
  (\citeauthor{li2015hiv}, \ensuremath{1.59\times 10^{3}} 
  [284 -
   \ensuremath{3.81\times 10^{3}}];
   \citeauthor{wang2015targeting}, 651
  [268 - 
   \ensuremath{4.24\times 10^{3}}]).
The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
estimated $I$, and HPD interval spanning almost the entire prior region
  (\ensuremath{7.55\times 10^{3}} 
  [228 -
   \ensuremath{8.92\times 10^{3}}]).
No information was gleaned about the $m$ parameter, with the HPD interval
occupying the entire prior region for all datasets. The estimates of $N$ were
similarly uninformative, with the exception that the point estimate for the
\citeauthor{wang2015targeting} data was smaller
  (\ensuremath{5.84\times 10^{3}})
than the estimates for other datasets
 (average \ensuremath{8.93\times 10^{3}}).

\begin{figure*}[ht]
  \centering
  \includegraphics{realdata-hpd-bc.pdf}
  \vspace{8pt}
  \caption{
      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
      parameters of the BA network model, fitted to five published HIV datasets
      with kernel-ABC.
  }
  \label{fig:abchpd}
\end{figure*}

>>>>

The average error for \gls{m} was
    1.31,
on a grid from 1 to 6; this error was positively correlated with increasing
\gls{m}
    (Spearman's $\rho$ = 0.25,
     $p = 6\!\times\!10^{-4}$).
This positive correlation was apparently due to the much lower error for
\gls{m} = 1 and 2 than for the other $m$ values 
    (mean errors 
    0.93 
    for $m \leq 2$ vs. 
    1.49 
    for $m > 2$, \cref{fig:gridm}).
The value $m = 1$ causes the network to take on a distinct shape relative to
higher \gls{m} values, namely a tree (\textit{i.e.} there are no cycles,
see \cref{subsec:treeshape}). The average error for \gls{N} was 
    2419,
on a grid from 1000 to 15000, and was positively correlated with the true value
of \gls{N}
    (Spearman's $\rho$ = 0.43,
    $p\ {<}10^{-5}$).

@@
